Tutorial for Runing LMD on Mouse Bone Marrow Dataset

### 2023-11-03 Demo_marrow_facs.Rmd
path adjustment

---
title: "Tutorial for Runing Mouse Bone Marrow Dataset"
author: "Ruiqi Li"
date: '2023-10-12'
output: pdf_document
bibliography: references.bib
---

# Install & Load Packages

```{r}
# Install necessary tools if not already present
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
# Define package lists
bioc_packages <- c("Seurat",
                   "clusterProfiler", "AnnotationDbi", "ReactomePA", "org.Mm.eg.db","gprofiler2")
github_packages <- c("satijalab/seurat-wrappers")
# Note: repeating packages from LMD_function.R as knitr does not install them properly
cran_packages <- c("igraph","ggplot2", "cowplot", "RColorBrewer",
                   "data.table", "dplyr", "patchwork",
                   "pheatmap", "ggplotify", "ggraph",
                   "ClusterR", "Rcpp", "RcppArmadillo", "tictoc")

# Install packages
sapply(cran_packages, function(pkg) if(!requireNamespace(pkg, quietly = TRUE)){install.packages(pkg)})
sapply(bioc_packages, function(pkg) if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg))
sapply(github_packages, function(pkg) if (!requireNamespace(strsplit(pkg, "/")[[1]][2], quietly = TRUE)) remotes::install_github(pkg))

library("Seurat")
library("SeuratWrappers")
library("pheatmap")
library("ggplot2")
library("ggplotify")
```

# Preparing input data

```{r}
if (!file.exists('facs_Marrow_seurat_tiss.Robj')) {
   # If load url doesn't work, please download it and load from your working directory
  options(timeout=6000)
  download.file("https://figshare.com/ndownloader/files/13092380",
                destfile = 'facs_Marrow_seurat_tiss.Robj', method = 'libcurl')
 }
load('facs_Marrow_seurat_tiss.Robj')
tiss <- UpdateSeuratObject(tiss)
```
We obtain the following objects from this Seurat Obj:
* `feature_space`: The output of PCA. This matrix contains the first 20 principal components (PCs).
* `visual_space`: a data.frame with t-SNE coordinates (2D) based on the first 20 PCs.
* `dat`: a matrix object with the log-normalized expression of genes (rows) in each cell (columns)
* `cell_label`: Any metadata you wish to visualize and examine related to cells (here we take the cell type annotation)
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class
```

@@

```{r}
source("./LMD_function.R", echo = F)
```

@@

# Denoising
#' We use ALRA @linderman2022zero to Denoise our data
tiss <- RunALRA(tiss, assay = "RNA")
DefaultAssay(tiss) <- "RNA"
dat_alra = as.matrix(tiss[["alra"]]@data)[local_gene,]

@@

## Step3: Identify Celltype-specific activated Modules
```{r}
cell_label = as.factor(tiss$cell_ontology_class)
abbrev = levels(cell_label)
abbrev[grepl("T cell",abbrev)] = "T cell"
abbrev[grepl("natural killer",abbrev)] = "NK cell"
abbrev[abbrev == "B cell"] = "Cd3e+ Klrb1+ B cell"
abbrev[grepl("Slamf1-positive",abbrev)] = "Slamf1+ MPC"
abbrev[grepl("Slamf1-negative",abbrev)] = "Slamf1- MPC"
abbrev[grepl("common lymphoid",abbrev)] = "CLP"
abbrev[grepl("megakaryocyte",abbrev)] = "MEP"
abbrev[grepl("hematopoietic",abbrev)] = "HSC"
abbrev[grepl("hematopoietic",abbrev)] = "HSC"
abbrev[grepl("granulocyte monocyte",abbrev)] = "GMP"
# abbrev[grepl("granulocytopoietic",abbrev)] = "G-poiesis"
names(abbrev) = levels(cell_label)
levels(cell_label) = abbrev[levels(cell_label)]

@@

# # Order Cell Type based on its centroid in Featurespace
# Idents(tiss) = cell_label
# ordered_cell_types <- unique(cell_label)[hclust(dist(t(sapply(unique(cell_label), function(x) colMeans(Embeddings(tiss, reduction = "pca")[WhichCells(tiss, idents = x), ,drop = FALSE]) ))))$order]

# Calculate Jaccard index between module activity & cell type
jaccard_index <- apply(cell_block, 2, function(module_col) {
  apply(one_hot_matrix, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard", mute.message=TRUE)
  })
})
# jaccard_index = jaccard_index[rev(ordered_cell_types),]
# rm(ordered_cell_types)
```

@@

## Transferability of Modules
### Load Data & Calculate Module Activity
```{r}
if (!file.exists('droplet_Marrow_seurat_tiss.Robj')) {
   # If load url doesn't work, please download it and load from your working directory
  options(timeout=6000)
  download.file("https://figshare.com/ndownloader/files/13092380",
                destfile = 'droplet_Marrow_seurat_tiss.Robj', method = 'libcurl')
 }
load('droplet_Marrow_seurat_tiss.Robj')
tiss_droplet <- UpdateSeuratObject(tiss_droplet)

# path <- "/data/ruiqi/local_marker/dataset/tabular_muris/"
# tissue_name = "marrow_droplet"
# tiss_droplet <- readRDS(sprintf("%s%s.rds",path,tissue_name))

@@

jaccard_index_droplet <- apply(cell_block_droplet, 2, function(module_col) {
  apply(one_hot_matrix_droplet, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard", mute.message=TRUE)
  })
})
```
## Save Intermediate data
```{r}
save(gene_partition,dist,gene_hree,cell_block,cell_block_droplet,one_hot_matrix,one_hot_matrix_droplet,jaccard_index,jaccard_index_droplet,file = "./marrow_result/module_result.rdata")
```

@@

# Set all genes(after filtering) in the expression matrix as background,
# such that the enrichment result of any module is not attributed to its high expression levels
universe_df = data.frame("symbol" = rownames(dat),"entrez" = mapIds(org.Mm.eg.db, keys=rownames(dat), column="ENTREZID", keytype="SYMBOL"))

@@

# Set all genes(after filtering) in the expression matrix as background,
# such that the enrichment result of any module is not attributed to its high expression levels
universe_df = data.frame("symbol" = rownames(dat),"entrez" = mapIds(org.Mm.eg.db, keys=rownames(dat), column="ENTREZID", keytype="SYMBOL"))
universe_df = universe_df[!is.na(universe_df$entrez),]

@@

## Step5: Generate Gene Module Annotation Table
```{r}
df_module = data.frame(Module = levels(gene_partition))
df_module[,"Gene"] = unlist(lapply(levels(gene_partition),function(x){paste(names(gene_partition)[gene_partition == x],collapse = ",")}))
df_module[,"celltype"] = module_celltype[levels(gene_partition)]
df_module[top_go_clusters,"GO_description"] = unlist(lapply(top_enrich_go,function(x){
  des = x[["Description"]]
  paste(des[!is.na(des)],collapse = ",")
}
))
df_module[top_pathway_clusters,"Pathway_description"] = unlist(lapply(top_enrich_pathway,function(x){
  des = x[["Description"]]
  paste(des[!is.na(des)],collapse = ",")
}))
df_module[is.na(df_module)] = ""
write.csv(df_module,file = "./marrow_result/module_description.csv",row.names = FALSE)
```

@@

write.csv(top_enrich_pathway_cc,file = "./marrow_result/reactome_pathway_cc.csv")
```
### Annotate Cell Cycle
```{r}
library(gprofiler2)
mmus_s = gorth(cc.genes.updated.2019$s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m = gorth(cc.genes.updated.2019$g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
tiss = CellCycleScoring(tiss, s.features = mmus_s, 
                        g2m.features = mmus_g2m, set.ident = FALSE)
cell_cycle_table = tiss@meta.data[,c("S.Score","G2M.Score","Phase")]
write.csv(cell_cycle_table,file = "./marrow_result/annotated_cell_cycle.csv")
```

<!-- ## Multi-resolution Modules - analysis -->
<!-- ```{r} -->
<!-- df_module = read.csv("./marrow_result/module_description.csv") -->
<!-- df_module$'multi-celltype' = NA -->
<!-- df_module$'note' = NA -->

<!-- ct_prob_per_module = do.call(rbind,lapply(unique(cell_label),function(x){apply(cell_block[cell_label == x,],2,median)})) -->
<!-- rownames(ct_prob_per_module) = unique(cell_label) -->
<!-- ct_num_per_module = sort(apply(ct_prob_per_module,2,function(x)sum(x > 0.5))) -->

<!-- m = 19 -->
<!-- df_module[m,"celltype"] = paste0(names(which(ct_prob_per_module[,m] > 0.5)),collapse = ",") -->

<!-- df_module[54,"multi-celltype"] = c("T/NK/NKT") -->
<!-- df_module[27,"multi-celltype"] = c("Progenitor/NK/NKT") -->
<!-- df_module[33,"multi-celltype"] = c("Progenitor/basophil/granulocytopoietic/half-granulocyte") -->
<!-- df_module[46,"multi-celltype"] = c("monocyte/granulocytopoietic/half-granulocyte") -->
<!-- df_module[48,"multi-celltype"] = c("monocyte/granulocytopoietic/granulocyte") -->
<!-- df_module[49,"multi-celltype"] = c("monocyte/macrophage/granulocytopoietic/granulocyte") -->
<!-- df_module[55,"multi-celltype"] = c("NK") -->
<!-- df_module[56,"multi-celltype"] = c("NK") -->
<!-- df_module[13,"multi-celltype"] = c("B") -->
<!-- df_module[50,"multi-celltype"] = c("monocyte/granulocytopoietic/granulocyte/basophil") -->
<!-- df_module[28,"multi-celltype"] = c("Progenitor") -->
<!-- df_module[9,"multi-celltype"] = c("immature/naive B") -->
<!-- df_module[11,"multi-celltype"] = c("late pro/pre B") -->
<!-- df_module[7,"multi-celltype"] = c("late pro/pre B") -->
<!-- df_module[44,"multi-celltype"] = c("granulocytopoietic/half-granulocyte") -->
<!-- df_module[47,"multi-celltype"] = c("granulocytopoietic/granulocyte") -->
<!-- df_module[52,"multi-celltype"] = c("T") -->
<!-- df_module[53,"multi-celltype"] = c("macrophage/NK") -->
<!-- df_module[1,"multi-celltype"] = c("monocyte/macrophage/granulocyte monocyte progenitor cell") -->
<!-- df_module[20,"multi-celltype"] = c("granulocytopoietic/Progenitor") -->
<!-- df_module[c(42,35),"multi-celltype"] = c("below-half-granulocyte/monocyte") -->
<!-- df_module[c(36,38),"multi-celltype"] = c("granulocyte monocyte progenitor cell/monocyte") -->
<!-- df_module[2,"multi-celltype"] = c("granulocyte monocyte progenitor cell/macrophage") -->
<!-- df_module[4,"celltype"] = c("granulocyte monocyte progenitor cell") -->
<!-- df_module[c(5,6,43),"note"] = "subtype" -->

<!-- df_module[c(16,18,17,31,32),'note'] = "cell_cycle" -->
<!-- ``` -->

>>>>

---
title: "Tutorial for Runing LMD on Mouse Bone Marrow Dataset"
author: "Ruiqi Li"
date: '2023-10-12'
output: pdf_document
bibliography: references.bib
---

# Install & Load Packages
```{r}
# Install necessary tools if not already present
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
# Define package lists
bioc_packages <- c("Seurat",
                   "clusterProfiler", "AnnotationDbi", "ReactomePA", "org.Mm.eg.db","gprofiler2")
github_packages <- c("satijalab/seurat-wrappers")
# Note: repeating packages from LMD_function.R as knitr does not install them properly
cran_packages <- c("igraph","ggplot2", "cowplot", "RColorBrewer",
                   "data.table", "dplyr", "patchwork",
                   "pheatmap", "ggplotify", "ggraph",
                   "ClusterR", "Rcpp", "RcppArmadillo", "tictoc")

# Install packages
sapply(cran_packages, function(pkg) if(!requireNamespace(pkg, quietly = TRUE)){install.packages(pkg)})
sapply(bioc_packages, function(pkg) if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg))
sapply(github_packages, function(pkg) if (!requireNamespace(strsplit(pkg, "/")[[1]][2], quietly = TRUE)) remotes::install_github(pkg))
library("Seurat")
library("SeuratWrappers")
library("pheatmap")
library("ggplot2")
library("ggplotify")
```
# Create a folder to save results
```{r}
folder_path = "marrow_result"
if (!dir.exists(path)) {
  # If it doesn't exist, create it
  dir.create(path)
} else {
}
```

# Preparing input data
```{r}
if (!file.exists('facs_Marrow_seurat_tiss.Robj')) {
   # If load url doesn't work, please download it and load from your working directory
  options(timeout=6000)
  download.file("https://figshare.com/ndownloader/files/13092380",
                destfile = 'facs_Marrow_seurat_tiss.Robj', method = 'libcurl')
 }
load('facs_Marrow_seurat_tiss.Robj')
tiss <- UpdateSeuratObject(tiss)
```
We obtain the following objects from this Seurat Obj:
* `feature_space`: The output of PCA. This matrix contains the first 20 principal components (PCs).
* `visual_space`: a data.frame with t-SNE coordinates (2D) based on the first 20 PCs.
* `dat`: a matrix object with the log-normalized expression of genes (rows) in each cell (columns)
* `cell_label`: Any metadata you wish to visualize and examine related to cells (here we take the cell type annotation)
```{r}
DefaultAssay(tiss) <- "RNA"
n_dim = dim(tiss@reductions$pca@cell.embeddings)[2]
feature_space = as.matrix(tiss@reductions$pca@cell.embeddings[,1:n_dim])
visual_space = data.frame(tiss@reductions$tsne@cell.embeddings)
dat = as.matrix(tiss[[DefaultAssay(tiss)]]@data)
cell_label = tiss$cell_ontology_class
```

@@

```{r}
source("LMD_function.R", echo = F)
```

@@

# Denoising
# We use ALRA @linderman2022zero to Denoise our data
tiss <- RunALRA(tiss, assay = "RNA")
DefaultAssay(tiss) <- "RNA"
dat_alra = as.matrix(tiss[["alra"]]@data)[local_gene,]

@@

## Step3: Identify Celltype-specific activated Modules
```{r}
cell_label = as.factor(tiss$cell_ontology_class)
abbrev = levels(cell_label)
abbrev[grepl("T cell",abbrev)] = "T cell"
abbrev[grepl("natural killer",abbrev)] = "NK cell"
abbrev[abbrev == "B cell"] = "Cd3e+ Klrb1+ B cell"
abbrev[grepl("Slamf1-positive",abbrev)] = "Slamf1+ MPC"
abbrev[grepl("Slamf1-negative",abbrev)] = "Slamf1- MPC"
abbrev[grepl("common lymphoid",abbrev)] = "CLP"
abbrev[grepl("megakaryocyte",abbrev)] = "MEP"
abbrev[grepl("hematopoietic",abbrev)] = "HSC"
abbrev[grepl("hematopoietic",abbrev)] = "HSC"
abbrev[grepl("granulocyte monocyte",abbrev)] = "GMP"
names(abbrev) = levels(cell_label)
levels(cell_label) = abbrev[levels(cell_label)]

@@

# Calculate Jaccard index between module activity & cell type
jaccard_index <- apply(cell_block, 2, function(module_col) {
  apply(one_hot_matrix, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard", mute.message=TRUE)
  })
})
```

@@

## Transferability of Modules
### Load Data & Calculate Module Activity
```{r}
if (!file.exists('droplet_Marrow_seurat_tiss.Robj')) {
   # If load url doesn't work, please download it and load from your working directory
  options(timeout=6000)
  download.file("https://figshare.com/ndownloader/files/13092380",
                destfile = 'droplet_Marrow_seurat_tiss.Robj', method = 'libcurl')
 }
load('droplet_Marrow_seurat_tiss.Robj')
tiss_droplet <- UpdateSeuratObject(tiss)
saveRDS(tiss_droplet, file = "processed_droplet_Marrow_srat.rds")

@@

jaccard_index_droplet <- apply(cell_block_droplet, 2, function(module_col) {
  apply(one_hot_matrix_droplet, 2, function(one_hot_col) {
    1 - philentropy::distance(rbind(module_col,one_hot_col), method = "jaccard", mute.message=TRUE)
  })
})
```
## Save Intermediate data
```{r}
save(gene_partition,dist,gene_hree,cell_block,cell_block_droplet,one_hot_matrix,one_hot_matrix_droplet,jaccard_index,jaccard_index_droplet,file = file.path(folder_path, "module_result.rdata"))

saveRDS(tiss, file = "processed_facs_Marrow_srat.rds")
```

@@

# Set all genes(after filtering) in the expression matrix as background
universe_df = data.frame("symbol" = rownames(dat),"entrez" = mapIds(org.Mm.eg.db, keys=rownames(dat), column="ENTREZID", keytype="SYMBOL"))

@@

# Set all genes(after filtering) in the expression matrix as background
universe_df = data.frame("symbol" = rownames(dat),"entrez" = mapIds(org.Mm.eg.db, keys=rownames(dat), column="ENTREZID", keytype="SYMBOL"))
universe_df = universe_df[!is.na(universe_df$entrez),]

@@

## Step5: Generate Gene Module Annotation Table
```{r}
df_module = data.frame(Module = levels(gene_partition))
df_module[,"Gene"] = unlist(lapply(levels(gene_partition),function(x){paste(names(gene_partition)[gene_partition == x],collapse = ",")}))
df_module[,"celltype"] = module_celltype[levels(gene_partition)]
df_module[top_go_clusters,"GO_description"] = unlist(lapply(top_enrich_go,function(x){
  des = x[["Description"]]
  paste(des[!is.na(des)],collapse = ",")
}
))
df_module[top_pathway_clusters,"Pathway_description"] = unlist(lapply(top_enrich_pathway,function(x){
  des = x[["Description"]]
  paste(des[!is.na(des)],collapse = ",")
}))
df_module[is.na(df_module)] = ""
write.csv(df_module,file = file.path(folder_path, "module_description.csv"),row.names = FALSE)
```

@@

write.csv(top_enrich_pathway_cc,file = file.path(folder_path, "reactome_pathway_cc.csv"))
```

### Annotate Cell Cycle
```{r}
mmus_s = gprofiler2::gorth(cc.genes.updated.2019$s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m = gprofiler2::gorth(cc.genes.updated.2019$g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
tiss = CellCycleScoring(tiss, s.features = mmus_s, 
                        g2m.features = mmus_g2m, set.ident = FALSE)
cell_cycle_table = data.frame(tiss@meta.data[,c("S.Score","G2M.Score","Phase")],"Dataset" = "FACS")
```

### Transferablity
```{r}
tiss_droplet = CellCycleScoring(tiss_droplet, s.features = mmus_s, 
                        g2m.features = mmus_g2m, set.ident = FALSE)
cell_cycle_table = rbind(cell_cycle_table,data.frame(tiss_droplet@meta.data[,c("S.Score","G2M.Score","Phase")],"Dataset" = "Droplet"))

write.csv(cell_cycle_table,file = file.path(folder_path, "annotated_cell_cycle.csv"))
```
